### 2016-01-27 intro.Rmd
ENH: More.

# Materials and Methods

## Imaging

## Quantitative analysis

<!--

* Importance of machine learning in medical image analysis
    x give examples
    x (tools) support vector machines
    x (tools) neural networks
    x (apps)
    x Random forests algorithms
        x MS lesion detection (Geremeria 2011)
        x Brain tumor segmentation (BRATS 2012 and BRATS 2013)
        x Tustison 2015 (importance of feature images)
        x Pustina 2016
    x Random forests overview

* Overview of protocol
    * Preprocess all images
        * Denoise
        * N4 bias correction
    * Create Stage 1 feature images
    * Use those as input to the first stage RF model to get RF voting maps
        * CSF
        * GM
        * WM
        * Deep GM
        * Brain stem
        * Cerebellum
        * WMHs
    * Use RF voting maps as priors for antsAtroposN4.sh to perform Bayesian-based
      segmentation with an additional Markov random field prior to promote spatial
      smoothness of the labels.
    * Use Stage 1 feature images + RF voting maps + antsAtroposN4.sh posteriors as
      Stage 2 feature images.
    * Remove isolated voxels as final refinement.

* Creation of feature images (all contained within ANTs)
    * Benefits of open source
    * best registration (for template building)
    * Preprocessing
        * Denoising
        * N4 bias correction
    * Types of feature images and motivation for their use (Stage 1)
        * Normalized intensities and neighborhood statistics image
        * Symmetric template --- the brain exhibits a shape and intensity bilateral
          symmetry under normal conditions.  WMHs can "break" this symmetry.
        * Probabilistic segmentations, distance maps and symmetric template indices ---
          help remove false positives
          caused by partial voluming at the gray/white matter interface and the location
          indices since, for example, higher intensities can be found at the periventricular
          caps in normal subjects [@Neema:2009aa] which often confounds automated lesion
          detection algorithms.
    * Types of feature images and motivation for their use (Stage 2)
        * All previously images (for the reasons stated above)
        * ``antsAtroposN4.sh`` with a large prior weight (w=0.5) on all three modalities
          to classify voxels based on intensity (use MRF to smooth out remaining noise
          in labeling).

* Public avail

* Leave one out evaluation strategy overview
    * Create stage 1 feature images for all subjects
    * For each of the 24 subjects, take out a subject and create a Stage 1 random forest
      model from the remaining feature images + manually labeled truth data.
    * For each of the remaining 23 subjects, send them through the Stage 1 RF model.
    * The above step permits the creation of the Stage 2 feature images
    * Create the Stage 2 RF model from the 23 sets of Stage 2 feature images.
    * Send the leave-one-out subject through the two RF models (as described above).
    * Compare with ground truth.

* (Save for Discusion) Difficulties of WMH evaluation (outlined in greater detail here [@MIAreview].
    *
-->

### Random forests for machine learning

@@

Crucial to these supervised segmentation approaches are the creation and selection of
"features" as input in conjunction with the ground-truth for model construction.  For the
targeted application in this work (i.e., white matter hyperintensities),
regression/classification are performed at the voxelwise level.  In other words, each voxel
within the region of interest is sent through the ensemble of decision trees and receives
a set of classification votes from each tree permitting a regression or classification
solution.  Since this procedure is performed at the voxelwise level, intensity information
alone is insufficient for good segmentation performance since it lacks spatial context.
For example, as pointed out in [@Neema:2009aa], higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated lesion detection
algorithms.  Other potential confounds include MR signal inhomogeneity and noise.  Therefore,
even though machine learning and pattern recognition techniques are extremely powerful and
have significant potential, just as crucial to outcome is the creative construction and
deployment of salient feature images which we detail below.

### Feature images for WMH segmentation

\input{featureImagesTable.tex}

Supervised methodologies are uniquely characterized, in part, by the feature images that
are used to identify the regions of interest.  In Table 1, we provide a list and basic
categorization of the feature images used for the initial (i.e., Stage 1---more on the use
of multiple random forest stages below) segmentation of the white matter hyperintensities.
In addition Figure 1 provides a representation of a set of feature images for a single
subject analyzed in this work.

![Representation of Stage 1 feature images for subject 01C1019.  The
FLAIR, T1-, and T2-weighted images are rigidly pre-aligned
[@Avants:2014aa] to the space of the T1 image.  The three images are then preprocessed
(N4 bias correction [@Tustison:2010ac] and adaptive denoising [@Manjon:2010aa]) followed by
application of standard ANTs brain extraction and $n$-tissue segmentation protocols using
the MMRR symmetric template corresponding and priors [@Tustison:2014ab] to the T1 image.
The feature images are then generated for voxelwise input to the RF model which results in
the voting maps illustrated on the right which gives a probabilistic classification of
tissue type.  Not shown are the probability and voting images for the brain stem and
cerebellum.](Figures/featureImages.png)

As mentioned previously, input for each subject comprises FLAIR, T1-, and T2-weighted
acquisitions.  The FLAIR and T2 images are rigidly registered to the T1 image using
the open-source Advanced Normalization Tools (ANTs) [@Avants:2014aa].  The aligned images
are then preprocessed using the denoising algorithm of [@Manjon:2010aa] followed by
N4 bias correction [@Tustison:2010ac] which are then normalized to the intensity
range $[0,1]$.  Although we could have used the intensity standardization algorithm of
[@nyul2000], we found that a simple linear rescaling produced better results.

The T1 image is then processed via the ANTs brain
extraction and tissue segmentation protocols described in [@Tustison:2014ab] in order to
produce a mask for the brain parenchyma and provide probabilistic estimates of the
cerebrospinal fluid (csf), gray matter, white matter, deep gray matter, brain stem, and
cerebellum.  Segmentation is performed using the ANTs Atropos tool [@Avants:2011aa] and
multi-model optimal symmetric shape/intensity templates [@Tustison:2014aa] created from the public
MMRR data set  [@landman2011].

To model the intensity information the first set of images simply includes the
preprocessed and normalized intensity FLAIR, T1, and T2 images.  We also calculate a set
of neighborhood statistics (mean, standard deviation, and skewness) feature images using
a radius of one voxel. For each of the normalized
images, we calculate the difference in intensities with the corresponding warped template
component.  Previous success in the international brain tumor segmentation
competition [@Menze:2015aa] was based on an important set of intensity features that were created
from multi-modal templates mentioned previously [@Tustison:2014aa].  We employ the
same strategy here.  For example, the template
difference feature image for the FLAIR image, $S_{FLAIR}$ is calculated as:

$$S_{FLAIR} - T_{FLAIR}\left(\phi_b^{-1}\right)$$

where $\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$ is the transform
which maps from the individual subject space to the
template space and $T_{FLAIR}$ is the FLAIR template component.  Also,
to take advantage of the bilateral symmetry of the normal brain (in terms of both shape
and intensity), and the fact that the presence of WMH violates that assumption,
we use the symmetric templates to compute the contralateral intensity
differences as an additional intensity feature.  For the FLAIR component, this contralateral
difference image is calculated from

$$S_{FLAIR} - S_{FLAIR}\left(\phi_b^{-1}\left(\phi_R\left(\phi_b\right)\right)\right)$$

where $\phi_R$ denotes a horizontal reflection perpendicular to the mid-sagittal plane of
the symmetric template.

The segmentation probability images described above are used as feature images to provide a
spatial context for the random forest model prediction step.  Additional spatial contextual
feature images include the distance maps [@maurer2003] based on the csf, gray matter, and
deep gray matter images.  These latter images are intended to help distinguish white matter
hyperintensities from false positives induced by the partial voluming at the gray/white
matter interface.  A third set of images are based on the voxel location within the
space of the template.  The T1 image of the subject is registered to the T1 template
component using a B-spline variant [@Tustison:2013ac] of the well-known ANTs Symmetric
Normalization (SyN) algorithm [@Avants:2011ab].  Since the inverse transform is also
derived as part of the registration process, we can warp the voxel index locations
back to the space of the individual subject.    Note that this is similar in motivation to the work
of [@Anbeek:2004aa].  However, this previous work lacks the normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.

### Stacked/cascaded/concatenated random forests for improved segmentation performance

It has been previously observed that

### Code and data availability

### Evaluation protocol

# Results

![Average \texttt{MeanDecreaseAccuracy} plots generated from the creation of all 24 random
forest models for both Stage 1 and Stage 2 during the leave-one-out evaluation.  These plots
are useful in providing a quantitative assessment of the predictive importance of each feature.
The error bars provide the $95^{th}$ percentile (i.e., $1.96 \times \sigma$) and illustrate the
stability of the feature importance across the leave-one-out models.](Figures/averageLeaveOneOut.png)

To calculate this quantity for a single feature from a single random forest model, the
decrease in prediction accuracy produced by omitting the specified feature is calculated
during the out-of-bag phase of model creation.

During the out-of-bag error calculation stage of the random forest model creation, the
decrease in prediction accuracy with the omission of a single feature or variable is
tracked and averaged. Those features which have the greatest decrease in mean accuracy are
considered to be the most discriminative. In this work, we do not use these measurements for
feature pruning.

\clearpage

# References

>>>>

# Introduction

## White matter hyperintensities in TBI

## Random forests for WMH segmentation

@@

In this work, we develop a concatenated random forest framework with a tailored feature image set
for segmenting white matter hyperintensities in traumatic brain injury cohorts.  Additionally,
the entire framework is provided as open source through the well-known open-source
ANTs[^i1] and ANTsR[^i2] toolkits.  Further motivating the investigation of this work is
the additional public availability of the TBI cohorts thus permitting full reproducibility
of the results reported and discussed.

[^i1]: https://github.com/stnava/ANTs
[^i2]: https://github.com/stnava/ANTsR

### 2016-01-27 methods.Rmd
ENH: More.

### Random forests for machine learning

Machine learning and pattern recognition techniques have seen increased application
for various medical image analysis workflows (see, for example, the annual
Workshop on Machine Learning in Medical Imaging held in conjunction with the Medical
Image Computing and Computer-Aided Intervention international meeting [@mlmi2015]).
Popular techniques such as support vector machines and neural networks have been applied
successfully to clinically relevant imaging tasks such as supervised image segmentation
(e.g., [@Bauer:2011aa]) and diagnostic prediction (e.g., [@Tong:2014aa;@Liu:2013aa]).
Facilitating the current employment of such techniques are the number of available imaging
data sets [@Van-Horn:2014aa] and the public availability of data science packages such as
SciPy [@scipy] and the R project for statistical computing [@R] and their associated add-on
toolkits.

Random forests [@breiman2001] is a popular machine learning technique that has
demonstrated significant utitility for supervised segmentation tasks (e.g.,
normal human brain segmentation) and other computer vision applications
(e.g., human gait detection [@viola2005]).
In the context of neuropathology, random forest-based paradigms have been employed in the
delineation of multiple sclerosis lesions [@geremia2011], stroke lesions [@Pustina:2016aa],
and brain tumors [@geremia2012;@bauer2012;@zikic2012;@Tustison:2014aa].  Of note, these
latter random forest approaches for brain tumor segmentation have performed well in recent
international competitions.  In response to the lack of objective comparisons between
segmentation algorithms, the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in
2012 [@Menze:2015aa] and has continued every year since under the auspices of the International Conference
of Medical Image Computing and Computer Assisted Interventions (MICCAI).

Random forests are conceptually simple [@breiman2001].  They consist of ensembles of decision trees
that are built from training data.  Once constructed, data to be classified is "pushed"
through each decision tree resulting in a single classification "vote" per tree.  These
votes are then be used for regression or classification of the data.  Although decision
trees had been extensively studied previously, the success of employing collections of
such weak learners for boosting machine learning performance
(e.g., AdaBoost [@schapire1990;@freund1997]) influenced the similarly sytled conglomeration
of decision trees into "forests" with randomized node optimization [@ho1995;@amit1997].
Finally, Breiman [@breiman2001] improved accuracy by random sampling of the training
data (i.e., "bagging") resulting in the current random forest framework.

@@

Supervised methodologies are uniquely characterized, in part, by the feature images that
are used to identify the regions of interest.  In Table 1, we provide a list and basic
categorization of the feature images used for the initial (i.e., Stage 1---more on the use
of multiple random forest stages below) segmentation of the white matter hyperintensities.
In addition Figure 1 provides a representation of a set of feature images for a single
subject analyzed in this work.

@@

The T1 image is then processed via the ANTs brain
extraction and tissue segmentation protocols described in [@Tustison:2014ab] in order to
produce a mask for the brain parenchyma and provide probabilistic estimates of the
cerebrospinal fluid (csf), gray matter, white matter, deep gray matter, brain stem, and
cerebellum.  Segmentation is performed using the ANTs Atropos tool [@Avants:2011aa] and
multi-model optimal symmetric shape/intensity templates [@Tustison:2014aa] created from the public
MMRR data set  [@landman2011].

@@

It has been previously observed that

@@

@@

# Results

![Average \texttt{MeanDecreaseAccuracy} plots generated from the creation of all 24 random
forest models for both Stage 1 and Stage 2 during the leave-one-out evaluation.  These plots
are useful in providing a quantitative assessment of the predictive importance of each feature.
The error bars provide the $95^{th}$ percentile (i.e., $1.96 \times \sigma$) and illustrate the
stability of the feature importance across the leave-one-out models.](Figures/averageLeaveOneOut.png)

To calculate this quantity for a single feature from a single random forest model, the
decrease in prediction accuracy produced by omitting the specified feature is calculated
during the out-of-bag phase of model creation.

During the out-of-bag error calculation stage of the random forest model creation, the
decrease in prediction accuracy with the omission of a single feature or variable is
tracked and averaged. Those features which have the greatest decrease in mean accuracy are
considered to be the most discriminative. In this work, we do not use these measurements for
feature pruning.

\clearpage

# References

>>>>

@@

Supervised methodologies are uniquely characterized, in part, by the feature images that
are used to identify the regions of interest.  In Table 1, we provide a list and basic
categorization of the feature images used for the initial (i.e., Stage 1---more on the use
of multiple random forest stages below) segmentation of the white matter hyperintensities.
In addition Figure 1 provides a representation of a set of feature images for a single
subject analyzed in this work.  Note that in this work we categorize the brain parenchyma
with seven labels:
* cerebrospinal fluid (label 1),
* gray matter (label 2),
* white matter (label 3),
* deep gray matter (label 4),
* brain stem (label 5),
* cerebellum (label 6), and
* white matter hyperintensities (label 7).

@@

The T1 image is then processed via the ANTs brain
extraction and tissue segmentation protocols described in [@Tustison:2014ab] in order to
produce a mask for the brain parenchyma and provide probabilistic estimates of the
cerebrospinal fluid (csf), gray matter, white matter, deep gray matter, brain stem, and
cerebellum.  These provide the ground-truth labels for the first six tissue labels
given above.  The white matter hyperintensities were manually identified by one of the authors
(J. R. S.) using the ITK-SNAP tool [@Yushkevich:2006aa].  Segmentation is performed using
the ANTs Atropos tool [@Avants:2011aa] and
multi-model optimal symmetric shape/intensity templates [@Tustison:2014aa] created from the public
MMRR data set  [@landman2011].

@@

In previous brain tumor segmentation work [@Tustison:2014aa], it was demonstrated that a
concatenated supervised approach, whereby the prediction output from the first random forest
model serves as partial input for a second random forest model, can significantly improve
segmentation performance.  We do the same thing for the work described here.  The Stage 1
feature images of the training data (as described previously) are used to construct the
Stage 1 model.  The training data Stage 1 features are then used to produce the voxelwise
voting maps via the Stage 1 model.  All the Stage 1 features plus the Stage 1 voting maps
are used as input to the Stage 2 model.  In addition, we use the Stage 1 voting maps as
tissue priors for a second application of the Atropos maximum aposteriori algorithm with an
additional Markov Random Field spatial prior (MAP-MRF) [@Avants:2011aa].  However, for
the second stage we use all three aligned preprocessed images for a multivariate
analysis.  The resulting seven posterior probability images constitute a third additional
feature image set for Stage 2.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion segmentation review
[@Garcia-Lorenzo:2013aa], although the number of algorithms reported in the literature
is quite extensive, there were only four publicly available segmentation algorithms
at the time of writing of which none are based on supervised learning.  As we did for
our brain tumor segmentation algorithm [@Tustison:2014aa], all of the code described in
this work is publicly available through the open-source ANTs/ANTsR toolkits.  In addition,
similar to our previous offering [^1], we plan on creating
a self-encapsulated example to showcase the proposed methodology.  The fact that this
the data will also be made available \textcolor{red}{through the FITBIR repository} along
with the ground truth data.

[^1]: https://github.com/ntustison/ANTsAndArboles

@@
